Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-09-27 VegOzone.tex
Remove sentences and add heyes reference

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to human health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone is a cause of visible injury, photosynthetic damage, early senescence as well as programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep[Chapter 2]{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions in relation to ozone impacts and air quality \citep{NCC:Lin2020}.

Complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic compounds (VOCs) in the presence of nitrogen oxides (\chem{NO_x}) lead to the formation of \chem{O_3} at ground-level. Despite a relatively short average tropospheric lifetime of approximately $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone and is precursors are subject to advection over long distances. Episodes of high ozone concentrations in the absence of local pollution sources can therefore often be attributed to the long-range transport of both ozone and its precursors. In northern Fennoscandia, episodes of enhanced ozone concentrations in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Besides VOCs from anthropogenic sources, hydrocarbons are also emitted by vegetation itself in the form of terpenes and monoterpenes, so-called biogenic volatile substances (BVOCs). These emissions are thought to be a response to factors, such as thermal stress, defense against herbivores, or even attraction of pollinators \citep{TPS:Penuelas2003}.

In the annual cycle in Boreal climates, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice-covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build-up of ozone and its precursors during winter. Come spring precursors are photochemically activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a  high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as described above as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

The start of the growing season (GS) in northern Fennoscandia has been shifting to earlier dates in response to the general warming trend \citep[e.g.]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby converging with the period of the ozone spring peak. At the same time, the growing season is also becoming longer. A longer growing season is prolonging the time in which vegetation can accumulate ozone. For natural and semi-natural vegetation which is already subject to rapid climatic changes known as Arctic amplification \citep{AMAP2012,IPCC2013}, these factors could promote a higher potential risk for northern vegetation to ozone-induced damage in the not so distant future.

>>>>

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to human health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone is a cause of visible injury, photosynthetic damage, early senescence as well as programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep[Chapter 2]{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions in relation to ozone impacts and air quality \citep{NCC:Lin2020}.

Complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic compounds (VOCs) in the presence of nitrogen oxides (\chem{NO_x}) lead to the formation of \chem{O_3} at ground-level. Despite a relatively short average tropospheric lifetime of approximately $22\,\unit{days}$ \citep{JGR:Stevenson2005,ACP:Young2013}, ozone and is precursors are subject to advection over long distances. Episodes of high ozone concentrations in the absence of local pollution sources can therefore often be attributed to the long-range transport of both ozone and its precursors. In northern Fennoscandia, episodes of enhanced ozone concentrations in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Besides VOCs from anthropogenic sources, hydrocarbons are also emitted by vegetation itself in the form of terpenes and monoterpenes, so-called biogenic volatile substances (BVOCs). These emissions are thought to be a response to factors, such as thermal stress, defense against herbivores, or even attraction of pollinators \citep{TPS:Penuelas2003}.

In the annual cycle in Boreal climates, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice-covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build-up of ozone and its precursors during winter. Come spring precursors are photochemically activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a  high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

The start of the growing season (GS) in northern Fennoscandia has been shifting to earlier dates in response to the general warming trend \citep[e.g.]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby converging with the period of the ozone spring peak. At the same time, the growing season is also becoming longer. A longer growing season is prolonging the time in which vegetation can accumulate ozone. For natural and semi-natural vegetation which is already subject to rapid climatic changes known as Arctic amplification \citep{AMAP2012,IPCC2013}, these factors could promote a higher potential risk for northern vegetation to ozone-induced damage in the not so distant future. Recently, \citet{ESPR:Hayes2021} showed that the highest sensitivity to future \chem{[O_3]} variability is expected in summer when vegetation is most productive.
